---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 3
)
```

# parameterpal

<!-- badges: start -->
<!-- badges: end -->

Rather than knowing intrinsically what parameters are required for a distribution, scientists tend to have a sense of what value they *expect* a measure to take, *how many* observations should fall *within* a certain distance of that value. 

For the normal distribution, this is straightforward, as the parameters reflect the expected value and variance. However, for the beta distribution, the parameters are not so readily interpretable. 

`parameterpal::` provides a means of obtaining the parameters required  for the beta distribution from interpretable conditions. 

## documentation

See this package's [`pkgdown::`](https://pkgdown.r-lib.org/articles/pkgdown.html)-generated site for more information.

## installation

```{r eval=FALSE}
devtools::install_github("softloud/parameterpal", build_vignettes = TRUE)
```

## intended user

This package was developed for a friend and collaborator, computational ecologist [Dr Matthew Grainger](https://github.com/DrMattG), who previously used browser-based tools to obtain beta parameters to inform his rstats workflow. I hope he finds this package a useful augment to his codeflow. 


# vignette

See `vignette("betapal")` for ~~more information~~ the same information, but from the handy ease of your Rstudio Viewer pane. 

Full disclosure, I need to update it after I 
fully understand this [gist](https://gist.github.com/daob/1422e978ff98bdf466fbcb4d9bf3e53e) I was provided with after posting the first version of this software to twitter. Open science at its best. Blogpost incoming on open science and why it's good to share bad math and beta research code. So cool I've ended up with a better, more mathematically correct solution after posting it. 

# other distributions

No reason the same math can't be applied to other distributions. Open an issue if you'd like me to provide parameters from interpretable conditions for another distribution.
