---
title: "Obtain beta parameters from interpretable conditions"
author: "Charles T. Gray"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{betapal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5
)

options(rmarkdown.html_vignette.check_title = FALSE)

```

```{r setup}
library(parameterpal)
library(ggplot2)

```


# motivation

Rather than knowing intrinsically what parameters are required for a distribution, scientists tend to have indirect knowledge. A researcher may have a  sense of what value they *expect* a measure to take, *how many* observations should fall *within* a certain distance of that value. As well as the overall shape of the data, the distribution. Distributions are not defined in the intuitive terms the scientist has, but equational parameters commonly (but not always) $\mu$ and $\sigma$ for the normal distribution, $\lambda$ for the exponential, and $\alpha$ and $\beta$ for the beta distribution. However, calculating the required parameters is not necessarily straightfoward, despite the coniditioning we receive from the normal distribution.

For the normal distribution, obtaining the parameters from these assmptions is straightforward. The expected value and variance, which is to say, centre and spread, of the normal distribution, are both easily interpretable and translate directly to the parameters required for the distribution. So,  to sample three values from a normal distribution where we expect a value of 0.5 with two-thirds of values falling within 0.2 of 0.5, we simply run the following code.

```{r}
rnorm(n = 3, 
      mean = 0.5,
      sd = 0.2)

```


But if we wished to sample from beta distribution, however, the parameters, `shape1` and `shape2`, are not readily interpretable from expected value and variance.  `parameterpal::` provides a means of obtaining the parameters required  for the beta distribution from interpretable conditions. 

## beta distribution

The beta distribution has some really nice properties. It is bounded by a minimum value of 0 and a maximum of 1, so is perfect for modelling proportions. 

Under some circumstances it can mimic a truncated bell curve, as well as flexibility for other shapes, such as a truncated parabola. 

But this is not intuitive. What shape do we expect $\mathrm{beta}(1,2)$ to take? It's hard to intuit from the parameters 1 and 2. But, we likely do not expect a negatively sloped line.

```{r}
ggplot() +
  xlim(0, 1) +
stat_function(
  fun = dbeta,
  args = list(shape1 = 1, shape2 = 2)
)



```


### original application

This code was developed for `softloud/simeta::` research software that supported `softloud`'s dissertation. 

In this case, given a population $N$, what proportion are allocated to the case and control groups? A desire to reflect the uncertainty of designed eXperiments motivated this code. 

For example, even if case and control groups were assignned evenly, in experimental science there are many reasons individuals may drop out of the groups. Thus a proportion of $N$ was sampled from a beta distribution (which is bounded between 0 and 1, as proportions are, too), with an intuition of how many drop outs are anticipated.

# using `parameterpal::` 

Suppose we expect a value of 0.3, with 80 per cent of observations falling within a distance of 0.2 from 0.3. That is, we expect 90 per cent of observations to fall within (0.1, 0.5). Assuming data follow a beta distribution, what are its parameters? 

```{r}
beta_pal(expected_value = 0.3,
         within = 0.2,
         this_much = 0.8)

```

We can plot this intuition to see the shape of the resulting beta distribution.

```{r }
beta_plot(expected_value = 0.3,
          within = 0.2,
          this_much = 0.8)

```

Escaping the ubiquitous tyranny of the arbitary bounds 90 or 95 per cent, which don't necessarily reflect our intuituion, suppose we thought only 30 per cent of values fall within the interval. 

```{r }
beta_plot(expected_value = 0.3,
          within = 0.2,
          this_much = 0.3)

```


# interpretability

The first parameter is the expected value, and the second, the variance. For the normal distribution, the standard deviation is interpretable. We know two-thirds of values fall within one standard deviation of the mean. If we shade this area in a visualisation, it's convincing that two-thirds of values fall within this range.

```{r}
expected_value <- 0.5
within <- 0.2

ggplot() +
  xlim(0, 1) +
  geom_rect(
    aes(
      xmin = expected_value - within,
      xmax = expected_value + within,
      ymin = 0,
      ymax = Inf,
      alpha = 0.2)) +
stat_function(
  fun = dnorm,
  args = list(mean = expected_value, sd = within)
) 


```

The beta distribution, on the other hand, requires two shape parameters, `shape1` and `shape2`, which do not immediately reflect our intuition of what value we expect the measure to take, nor how much variance we expect.

